### 2020-09-17 latex/vnhcs.tex
Edited phrasing up to and including simply actuated hamiltonian

\section{Preliminaries on Analytical Mechanics}
A mechanical system can be represented by \(N\) point masses where each point
represents the center of mass of a physical body, along with \(r\)
\textit{equations of constraint} (EOC) which model the physical restrictions
between these masses.
The position of each point mass is described using three cartesian coordinates (one
for each spatial axis), so the system as a whole can be described by a vector in
\(\R^{3N}\) with \(r\) EOC. 
The dynamics of the system are computed by deriving the \(3N\)
\textit{equations of motion} (EOM) produced by Newton's second law \(F = m a\).
While this technique works for simple systems, it is impossible to
apply to a majority of mechanical systems since most of the forces 
on the system are not explicitly known.

Rather than modeling a mechanical system by cartesion positions and constraints,
it is often feasible to represent the position of the system using \(n\)
independent scalar-valued variables \(q_1,\ldots,q_n\) called 
\textit{generalized coordinates}, where \(n = 3N - r\) is the number of
\textit{degrees of freedom} (DOF) of the system \cite{greenwood_dynamics}.

% TODO:
\textbf{TODO: Image comparing Newtonian to g.c.}

@@

Lagrangian mechanics uses the kinetic energy \(T(q,\dot{q})\) and potential
energy \(P(q)\) of the system to define the Lagrangian 
\(\mathcal{L} : \mathcal{Q}\times\R^n \rightarrow \R\) defined by
(\ref{eqn:lagrangian-general}) \cite{greenwood_dynamics}.
\begin{equation}\label{eqn:lagrangian-general}
    \mathcal{L}(q,\dot{q}) = T(q,\dot{q}) - P(q)
\end{equation}
When the mechanical system is actuated, the EOM are described by \(n\) second-order
ordinary differential equations (ODEs) obtained from the \textit{Euler-Lagrange
equations} (\ref{eqn:el-eqns-general}) with \textit{generalized input forces} 
\(\tau \in \R^k\) . 
\begin{equation}\label{eqn:el-eqns-general}
    \diff{}{t}\left\{ \pdiff{\mathcal{L}}{\dot{q}_i} \right\}
    - \pdiff{\mathcal{L}}{q_i} = B_i\tpose(q) \tau
\end{equation}
The vector \(B_i\tpose: \mathcal{Q} \rightarrow \R^{1\times k}\) describes how
the input forces shape the dynamics of \(q_i\).
The matrix  \(B: \mathcal{Q} \rightarrow \R^{n \times k}\) with
\[
    B(q) = \begin{bmatrix}
        - & B_1\tpose(q) & - \\
          & \vdots & \\
        - & B_n\tpose(q) & - \\
    \end{bmatrix}
\]
is called the \textit{input matrix} for the system.
If \(k < n\), we say the system is \textit{underactuated} with degree of
underactuation \(n - k\).

@@

% ---------- Hamiltonian Mechanics ---------- %
\subsection{Hamiltonian Mechanics}
Hamiltonian mechanics provides an equivalent representation of the EOM
by converting the \(n\) second-order ODEs generated by Lagrangian mechanics into
\(2n\) first-order ODEs.

To do this, we first define the \textit{conjugate of momentum \(p_i\) to \(q_i\)} by
\begin{equation}\label{eqn:p-i}
    p_i(q,\dot{q}) = \pdiff{\mathcal{L}}{\dot{q}_i}(q,\dot{q})
\end{equation}
To ease notation, we will write \(p = (p_1, \ldots, p_n) \in \R^n\) and say that
\(p\) is the \textit{conjugate of momenta to \(q\)}.
Note that each \(p_i\) is a linear function of \(\dot{q}\), and one can typically
solve for \(\dot{q}(q,p)\) by inverting all the expressions from (\ref{eqn:p-i}).
The combined vector \((q,p) \in \mathcal{Q}\times\R^n\) is called a 
\textit{phase} of the system.

The \textit{Hamiltonian} of the system in
\(\{q,p\}\) coordinates is defined as the Legendre transform 
(\ref{eqn:hamiltonian-legendre}) of the Lagrangian \cite{landau_mechanics}.
\begin{equation}\label{eqn:hamiltonian-legendre}
    \mathcal{H}(q,p) := p\tpose \dot{q}(q,p) - \mathcal{L}(q,\dot{q}(q,p))
\end{equation}
The EOM in this framework can be shown to be the \(2n\)
first-order equations called \textit{Hamilton's equations}
\begin{equation}\label{eqn:hamiltons-eqns}
    \begin{cases}
        \dot{q} = \nabla_p\mathcal{H} \\
        \dot{p} = -\nabla_q\mathcal{H} + B(q)\tau \\
    \end{cases}
\end{equation}
where \(B(q) \in \R^{n\times k}\) is the same input matrix used by the
Lagrangian and \(\tau \in \R^k\) is the vector of generalized input forces.

If the Lagrangian has quadratic kinetic energy as in (\ref{eqn:lagrangian}), 
the conjugate of
momenta to \(q\) can be computed explicitly by \(p = D(q)\dot{q}\).

The resulting Hamiltonian system reduces to (\ref{eqn:hamiltonian}).
\begin{align}\label{eqn:hamiltonian}
    \mathcal{H}(q,p) &= \frac{1}{2} p\tpose D^{-1}(q) p + P(q) \\
                     &\begin{cases}
        \dot{q} = D\inv(q)p \\
        \dot{p} = -\frac{1}{2} (\Id{n} \otimes p\tpose) \nabla_q D\inv(q) p
        - \nabla_q P(q) + B(q) \tau \\
    \end{cases} 
\end{align}

Any set of coordinates \(\{q,p\}\) which satisfy Hamilton's equations 
under the Hamiltonian \(\mathcal{H}\) are
said to be \textit{canonical coordinates} for the system. A change of
coordinates \((q,p) \rightarrow (Q,P)\) is a \textit{canonical
transformation} if \(\{Q,P\}\) are canonical coordinates with Hamiltonian
\(H(Q,P) = \mathcal{H}\left(q(Q,P), p(Q,P)\right)\).

\section{Simply Actuated Hamiltonian Systems}
% Describe the change of coordinates to get into q_u/q_a mode and show
% that the actuator directly affects pa but not pu. Use M for simply actuated
% inertia, D for normal coordinates
Given a Hamiltonian mechanical system (\ref{eqn:hamiltonian}), it is not obvious
how any particular input force \(\tau_i\) affects the conjugate of momenta. 
This is because \(\tau\) is transformed by the input matrix \(B(q)\) before
entering the EOM.

We will define a new class of Hamiltonian systems where the effect of the input
forces on the conjugate of momenta is made obvious. This class of systems will
form the backbone for the rest of the theory developed in this thesis.

\begin{defn}
    Let \(\mathcal{H}\) be an underactuated Hamiltonian system 
    \(\mathcal{H}\) with degree of underactuation \((n-k)\geq 0\). 
    A pair of canonical coordinates \(\{q,p\}\) for this system
    are said to be \textit{simply actuated coordinates} if the
    input matrix \(B(q) \in \R^{n \times k}\) is of the form
    \[
        B(q) = \simpleB    
    \]
    The first \((n-k)\) coordinates \(q_u\) are called the \textit{unactuated
    coordinates}, while the remaining \(k\) coordinates \(q_a\) are called the
    \textit{actuated coordinates}. We write \(q = (q_u, q_a)\); the
    corresponding conjugate of momenta \(p = (p_u, p_a)\) are called the
    unactuated and actuated momenta (respectively).
\end{defn}
\begin{defn}
    A Hamiltonian system is said to be \textit{simply actuated} if there exists
    a canonical transformation from any set of canonical coordinates 
    \(\{q,p\}\) into simply actuated coordinates for \(\mathcal{H}\).
\end{defn}

@@

\begin{assm}\label{assm:B-const}
    The input matrix \(B(q) \equiv B \in \R^{n\times k}\) is constant,
    full rank, and \(k \leq n\).
\end{assm}
\begin{assm}\label{assm:B-perp}
    There exists a matrix 
    \(B^\perp \in \R^{(n-k)\times n}\)
    which is right semi-orthogonal 
    (\(B^\perp(B^\perp)\tpose = \Id{(n-k)}\))
    and which is a left-annihilator for \(B\)
    (\(B^\perp B = \Zmat{(n-k) \times k}\)).
\end{assm}

Note that if \(k = (n-1)\), the existence of any left annihilator \(A^0\) for
\(B\) implies the left annihilator \(B^\perp := A^0/\norm{A^0}\) satisfies Assumption
\ref{assm:B-perp}.

\begin{assm}\label{assm:B-orthogonal}
    Assume \textbf{without loss of generality} that the input matrix \(B\) is
    left semi-orthogonal (\(B\tpose B = \Id{k}\)).
\end{assm}
\begin{proof}
Since \(B\) is a constant matrix, 
it has a singular-value decomposition 
\(B = U \Sigma V\tpose\) where \(U^{-1} = U\tpose \in \R^{n \times n}\), 
\(V^{-1} = V\tpose \in \R^{k \times k}\), and \(\Sigma \in \R^{n \times k}\) is
defined by
\[
    \Sigma = \begin{bmatrix}
        \sigma_1 & 0 & \cdots & 0 \\
        0 & \sigma_2 & \cdots & 0 \\
        \vdots & & \ddots & \vdots \\
        0 & 0 & \cdots & \sigma_k \\
        - &   & \Zmat{(n-k)\times k} & -  \\
    \end{bmatrix}
\]
% TODO: Find a source for SVD
where \(\sigma_i \neq 0\) because \(B\) is full-rank \textbf{TODO: source?}.
Defining \(T \in \R^{k \times k}\) by
\[
    T = \begin{bmatrix}
        \frac{1}{\sigma_1} & 0 & \cdots & 0 \\
        0 & \frac{1}{\sigma_2} & \cdots & 0 \\
    \vdots & & \ddots & \vdots \\
    0 & 0 & \cdots & \frac{1}{\sigma_k} \\
    \end{bmatrix}
\]
and assigning the input forces to \(\tau = V T \hat{\tau}\), we get a new input
matrix for \(\hat{\tau} \in \R^k\) given by \(\hat{B} = B V T = U \Sigma T\) 
which is still constant and full-rank. In particular, 
\(\hat{B}\tpose \hat{B} = T\tpose \Sigma\tpose \Sigma\tpose T = \Id{k}\).
\end{proof}

Let \(\mathbf{B} \in \R^{n\times n}\) be the following matrix:
\[
    \mathbf{B} = 
    \begin{bmatrix}
        B^\perp \\
        B\tpose \\
    \end{bmatrix}
\]
Since \(B^\perp\) is a left annihilator of \(B\) and both \(B^\perp\) and
\(B\tpose\) are right semi-orthogonal, it is easy to show that \(\mathbf{B}\) is
orthogonal:
\[
    \mathbf{B}\mathbf{B}\tpose = 
    \begin{bmatrix}
        B^\perp (B^\perp)\tpose & B^\perp B \\
        (B^\perp B)\tpose & B\tpose B
    \end{bmatrix} = \Id{n}
    \Rightarrow
    \mathbf{B}^{-1} = \mathbf{B}\tpose
\]

The following theorem shows that \(\mathbf{B}\) provides a canonical
transformation into simply actuated coordinates, so that all unactuated momenta
are unaffected by the input forces.

\begin{thm}\label{thm:simply-actuated}
    Under Assumptions \ref{assm:B-const},\ref{assm:B-perp}, and
    \ref{assm:B-orthogonal}, the Hamiltonian system (\ref{eqn:hamiltonian}) is
    simply actuated with simply actuated coordinates 
    \(\{Q = \mathbf{B}q, P = \mathbf{B}p\}\). The resulting dynamics are 
    given by (\ref{eqn:simple-hamiltonian}),
    \begin{align}\label{eqn:simple-hamiltonian}
        \hat{H}(Q,P) &= 
        \frac{1}{2} P\tpose \Minv(Q) P + V(Q) \\
       &\begin{cases}
            \dot{Q} = \Minv(Q)P \\
            \dot{P} = -\frac{1}{2} (\Id{n} \otimes P\tpose) \nabla_Q \Minv(Q) P
                - \nabla_Q V(Q) + \simpleB \tau
        \end{cases} \nonumber
    \end{align}
    where
    \begin{align*}
        \Minv(Q) &:= \mathbf{B}D^{-1}(\mathbf{B}\tpose Q)\mathbf{B}\tpose \\
        V(Q) &:= P(\mathbf{B}\tpose Q) \\
    \end{align*}
\end{thm}
\begin{proof}
    The Poisson bracket between the functions \(f(q,p)\) and \(g(q,p)\) is defined by
    \cite{landau_mechanics} as follows:
    \[
        [f,g] := \sum \limits_{i=1}^n \pdiff{f}{p_i}\pdiff{g}{q_i} - 
                \pdiff{f}{q_i}\pdiff{g}{p_i}
    \]
    For any constant matrix \(A\), the transformation
    \(\{Q = Aq, P = Ap\}\) satisfies
    \(\pdiff{Q_i}{p_m} = \pdiff{P_i}{q_m} = 0\) for all 
    \(i,m \in \n\).
    Hence, for our new coordinates \(\{Q = \mathbf{B}q, P = \mathbf{B}p\}\),
    \begin{align*}
        [Q_i,Q_j] &:= \sum\limits_{m = 1}^n \pdiff{Q_i}{p_m}\pdiff{Q_j}{q_m} - 
        \pdiff{Q_i}{q_m}\pdiff{Q_j}{p_m} = 0 \\
        [P_i,P_j] &:= \sum\limits_{m=1}^n \pdiff{P_i}{p_m}\pdiff{P_j}{q_m} -
        \pdiff{P_i}{q_m}\pdiff{P_j}{p_m} = 0
    \end{align*}
    Since the matrix \(\mathbf{B}\) is orthogonal,
    \((\mathbf{B}_i)\tpose (\mathbf{B}\tpose)_j = (\mathbf{B}_i)\tpose (\mathbf{B}^{-1})_j = \delta_{i,j}\). Using this
    fact we see that the Poisson brackets between \(P_i\) and \(Q_j\) are given by:
    \begin{align*}
        [P_i,Q_j] &= \sum\limits_{m=1}^n\pdiff{P_i}{p_m}\pdiff{Q_j}{q_m}
        - \pdiff{P_i}{q_m}\pdiff{Q_j}{p_m} \\
                  &= \sum\limits_{m=1}^n \mathbf{B}_{i,m}\mathbf{B}_{j,m} - 0 \\
                  &= \sum\limits_{m=1}^n \mathbf{B}_{i,m}\mathbf{B}\tpose_{m,j} \\
                  &= (\mathbf{B}_i)\tpose \mathbf{B}\tpose_j \\
                  &= \delta_{i,j}
    \end{align*}

    Therefore, by (45.10) in \cite{landau_mechanics}, the coordinate change 
    \((Q = \mathbf{B}q, P = \mathbf{B}p)\) is a canonical transformation with
    new Hamiltonian 
    \(\hat{H}(Q,P) = \mathcal{H}(\mathbf{B}\tpose Q,\mathbf{B}\tpose P)\).

    Furthermore, since \(\dot{P} = \mathbf{B}\dot{p}\), the new input matrix is
    given by 
    \[
        \mathbf{B}B = \begin{bmatrix}
            B^\perp B \\
            B\tpose B \\
        \end{bmatrix} = 
        \begin{bmatrix}
            \Zmat{(n-k)\times k} \\
            \Id{k}
        \end{bmatrix}
    \]
    so the coordinates \(\{Q = (q_u,q_a), P = (p_u,p_a)\}\) are simply actuated coordinates for 
    \(\hat{H}\) as desired.
\end{proof}

@@

The downside to VHCs is that they do not use the additional information imparted
by the generalized velocities of a system.
Many authors have attempted to extend VHCs to enforce these relations 
\(h(q,\dot{q}) = 0\) of the full state to account for this drawback. 
Since these relations use actuators to
restrict both the configuration and generalized velocity, they are called
virtual \textit{nonholonomic} constraints. This idea has been used for
human-robot interaction
\cite{vnhc_human_robot_cooperation,psd_based_vnhc_redundant_manipulator,haptic_vnhc},
error-reduction on time-delayed systems \cite{vnhc_time_delay_teleop},
and has shown marked improvements to the field of bipedal locomotion 
\cite{nhvc_dynamic_walking,
hybrid_zero_dynamics_bipedal_nhvcs,output_nhvc_bipedal_control}.
Most interestingly, the nonholonomic approach has been shown to be more robust
than standard VHCs when applied to bipedal robotics \cite{nhvc_incline_walking}.
This suggests that virtual nonholonomic constraints are more capable of
stabilizing specific energy levels while rejecting disturbances, all while
producing realistic biological motion. It is for this reason we choose to study
nonholonomic constraints in this thesis.

@@

    Moving on to \(\diff{dh_{p_u}}{t}\), a similar approach as

@@

    Now observe that the \(i^\text{th}\) row of

@@

    This allows us to write the entire derivative in matrix form:

>>>>

\section{Preliminaries on Analytical Mechanics}
A mechanical system can be represented by \(N\) point masses where each point
represents the center of mass of a physical body, along with \(r\)
\textit{equations of constraint} (EOC) which model the physical restrictions
between these masses.
The position of each point mass is described using three cartesian coordinates (one
for each spatial axis), so the system as a whole can be described by a vector in
\(\R^{3N}\) with \(r\) EOC. 
The dynamics of the system are computed by deriving the \(3N\)
\textit{equations of motion} (EOM) produced by Newton's second law \(F = m a\).
While this technique works for simple systems, it is tedius and becomes
impossible to apply to complex mechanical systems where the forces are not
explicitly known.

Rather than modeling a mechanical system by cartesian positions and constraints,
it is often feasible to represent the position of the system using \(n\)
independent scalar-valued variables \(q_1,\ldots,q_n\) called 
\textit{generalized coordinates}, where \(n = 3N - r\) is the number of
\textit{degrees of freedom} (DOF) of the system \cite{greenwood_dynamics}. For
instance, Figure \ref{fig:barbell} shows a barbell on a \(2D\)-plane which can
rotate freely on that plane. 
The barbell has \(n = 3\) DOF, so it can be described by three
independent generalized coordinates with no equations of constraint.

\begin{figure}
   \centering
   \begin{subfigure}[]{0.75\textwidth}
      \includestandalone[width=\textwidth]{images/newtonian_barbell}
      \caption{The Newtonian representation of the barbell 
      requires all six cartesian positions and the corresponding EOC.}
   \end{subfigure}
   \begin{subfigure}[]{0.75\textwidth}
      \includestandalone[width=\textwidth]{images/gc_barbell}
      \caption{One possible set of three generalized coordinates is
       \((x_0,y_0,\theta)\), which represent the position of the 
       center of the bar and the angle of the barbell in the \(xy\)-plane.}
   \end{subfigure}
   \caption{A mechanical system with \(N = 2\) points at \((x_1,y_1,z_1)\)
   and \((x_2,y_2,z_2)\) separated by a bar with \(r = 3\) EOC given by 
   \(c = \norm{x_1 - x_2}\), \(z_1 = 0\), and \(z_2 = 0\). This system has \(n =
   3\) degrees of freedom.}
   \label{fig:barbell}
\end{figure}

@@

Lagrangian mechanics uses the kinetic energy \(T(q,\dot{q})\) and potential
energy \(P(q)\) of the system to define the Lagrangian 
\(\mathcal{L} : \mathcal{Q}\times\R^n \rightarrow \R\) defined by
(\ref{eqn:lagrangian-general}) \cite{greenwood_dynamics}.
\begin{equation}\label{eqn:lagrangian-general}
    \mathcal{L}(q,\dot{q}) = T(q,\dot{q}) - P(q)
\end{equation}
When the mechanical system is actuated, the EOM are described by \(n\) second-order
ordinary differential equations (ODEs) obtained from the \textit{Euler-Lagrange
equations} (\ref{eqn:el-eqns-general}) with \textit{generalized input forces} 
\(\tau \in \R^k\) . 
\begin{equation}\label{eqn:el-eqns-general}
    \diff{}{t}\left\{ \pdiff{\mathcal{L}}{\dot{q}_i} \right\}
    - \pdiff{\mathcal{L}}{q_i} = B_i\tpose(q) \tau
\end{equation}
The vector \(B_i\tpose: \mathcal{Q} \rightarrow \R^{1\times k}\) describes how
the input forces shape the dynamics of \(q_i\).
The matrix  \(B: \mathcal{Q} \rightarrow \R^{n \times k}\) with
\[
    B(q) = \begin{bmatrix}
        - & B_1\tpose(q) & - \\
          & \vdots & \\
        - & B_n\tpose(q) & - \\
    \end{bmatrix}
\]
is called the \textit{input matrix} for the system.
If \(k < n\), we say the system is \textit{underactuated} with degree of
underactuation \((n - k)\).

@@

% ---------- Hamiltonian Mechanics ---------- %
\subsection{Hamiltonian Mechanics}
Hamiltonian mechanics converts the \(n\) second-order ODEs generated by
Lagrangian mechanics into an equivalent set of \(2n\) first-order ODEs.

To do this, we first define the \textit{conjugate of momentum \(p_i\) to \(q_i\)} by
\begin{equation}\label{eqn:p-i}
    p_i(q,\dot{q}) := \pdiff{\mathcal{L}}{\dot{q}_i}(q,\dot{q})
\end{equation}
To ease notation, we write \(p = (p_1, \ldots, p_n) \in \R^n\) and call
\(p\) the \textit{conjugate of momenta to \(q\)}.
Note that each \(p_i\) is a linear function of \(\dot{q}\), and one can typically
solve for \(\dot{q}(q,p)\) by inverting all the expressions from (\ref{eqn:p-i}).
The combined vector \((q,p) \in \mathcal{Q}\times\R^n\) is called a 
\textit{phase} of the system.

The \textit{Hamiltonian} of the system in
\(\{q,p\}\) coordinates is the ``Legendre transform"
(\ref{eqn:hamiltonian-legendre}) of the Lagrangian \cite{landau_mechanics}.
\begin{equation}\label{eqn:hamiltonian-legendre}
    \mathcal{H}(q,p) := p\tpose \dot{q}(q,p) - \mathcal{L}(q,\dot{q}(q,p))
\end{equation}
The EOM in the Hamiltonian framework are the \(2n\)
first-order equations called \textit{Hamilton's equations}. They are given by:
\begin{equation}\label{eqn:hamiltons-eqns}
    \begin{cases}
        \dot{q} = \nabla_p\mathcal{H} \\
        \dot{p} = -\nabla_q\mathcal{H} + B(q)\tau \\
    \end{cases}
\end{equation}
Here, \(B(q) \in \R^{n\times k}\) is the same input matrix used by the
Lagrangian framework, with \(\tau \in \R^k\) the same vector of generalized
input forces.

If the kinetic energy of the system is quadratic as in (\ref{eqn:lagrangian}), 
the conjugate of momenta becomes \(p = D(q)\dot{q}\). Since \(D(q)\) is symmetric
and positive definite, it is invertible at each \(q \in \mathcal{Q}\).
The resulting Hamiltonian system reduces to (\ref{eqn:hamiltonian}).
\begin{align}\label{eqn:hamiltonian}
    \mathcal{H}(q,p) &= \frac{1}{2} p\tpose D^{-1}(q) p + P(q) \\
                     &\begin{cases}
        \dot{q} = D\inv(q)p \\
        \dot{p} = -\frac{1}{2} (\Id{n} \otimes p\tpose) \nabla_q D\inv(q) p
        - \nabla_q P(q) + B(q) \tau \\
    \end{cases} 
\end{align}

Any set of coordinates \(\{q,p\}\) which satisfy Hamilton's equations 
under the Hamiltonian \(\mathcal{H}\) are
said to be \textit{canonical coordinates} for the system. A change of
coordinates \((q,p) \rightarrow (Q,P)\) is a \textit{canonical
transformation} if \(\{Q,P\}\) preserve the Hamiltonian structure; that is, if
they are canonical coordinates under the Hamiltonian
\(\mathcal{H}\left(q(Q,P), p(Q,P)\right)\).

\section{Simply Actuated Hamiltonian Systems}
\textbf{Manfredi: What term should we use instead of ``simply actuated"?}

% Describe the change of coordinates to get into q_u/q_a mode and show
% that the actuator directly affects pa but not pu. Use M for simply actuated
% inertia, D for normal coordinates
Suppose we are given a Hamiltonian mechanical system (\ref{eqn:hamiltonian}).
Because \(\tau\) is transformed by the input matrix \(B(q)\) before
entering the EOM, it is not in general clear how any particular input force \(\tau_i\)
will affect the dynamics of the system. 
In this section, we define a new class of Hamiltonian systems where the effect
of the input forces is made obvious. This class of
systems will form the backbone for the rest of the theory developed in this
thesis.

\begin{defn}
    Let \(\mathcal{H}\) be an \(n\)-DOF Hamiltonian system 
    with \(k \leq n\) actuators. 
    A set of canonical coordinates \(\{q,p\}\) for this system
    are said to be \textit{simply actuated coordinates} if the
    input matrix \(B(q) \in \R^{n \times k}\) is of the form
    \[
        B(q) = \simpleB    
    \]
    The first \((n-k)\) coordinates, labelled \(q_u\), are called the
    \textit{unactuated coordinates}. The remaining \(k\) coordinates, labelled
    \(q_a\), are called the \textit{actuated coordinates}. When grouping them
    together, we will always put them in the order \((q_u, q_a)\) to fit with 
    the definition. 
    The corresponding \((p_u, p_a)\) are called the unactuated and actuated
    momenta, respectively.
\end{defn}
\begin{defn}
    A Hamiltonian system is said to be \textit{simply actuated} if there exists
    a canonical transformation from any canonical coordinates 
    \(\{q,p\}\) into simply actuated coordinates.
\end{defn}

@@

\begin{assm}\label{assm:B-const}
    The input matrix \(B(q) \equiv B \in \R^{n\times k}\) is constant,
    full rank, and \(k < n\).
\end{assm}
\begin{assm}\label{assm:B-perp}
    There exists a matrix 
    \(B^\perp \in \R^{(n-k)\times n}\)
    which is right semi-orthogonal 
    \(\left(B^\perp(B^\perp)\tpose = \Id{(n-k)}\right)\)
    and which is a left-annihilator for \(B\). 
    That is, \(B^\perp B = \Zmat{(n-k) \times k}\).
\end{assm}

Note that if \(k = (n-1)\), the existence of any left annihilator 
\(A^0 \in \R^{1\times n}\) implies the left annihilator 
\(B^\perp := A^0/\norm{A^0}\) satisfies Assumption \ref{assm:B-perp}.

\begin{assm}\label{assm:B-orthogonal}
    Assume \textbf{without loss of generality} that the input matrix \(B\) is
    left semi-orthogonal. That is, \(B\tpose B = \Id{k}\).
\end{assm}
\begin{proof}
Since \(B\) is a constant matrix, 
it has a singular-value decomposition 
\(B = U \Sigma V\tpose\) where \(U^{-1} = U\tpose \in \R^{n \times n}\), 
\(V^{-1} = V\tpose \in \R^{k \times k}\), and \(\Sigma \in \R^{n \times k}\) is
defined by
\[
    \Sigma = \begin{bmatrix}
        \sigma_1 & 0 & \cdots & 0 \\
        0 & \sigma_2 & \cdots & 0 \\
        \vdots & & \ddots & \vdots \\
        0 & 0 & \cdots & \sigma_k \\
        - &   & \Zmat{(n-k)\times k} & -  \\
    \end{bmatrix}
\]
where \(\sigma_i \neq 0\) because \(B\) is full-rank \cite{calculating_svd}.
Defining \(T \in \R^{k \times k}\) by
\[
    T = \begin{bmatrix}
        \frac{1}{\sigma_1} & 0 & \cdots & 0 \\
        0 & \frac{1}{\sigma_2} & \cdots & 0 \\
    \vdots & & \ddots & \vdots \\
    0 & 0 & \cdots & \frac{1}{\sigma_k} \\
    \end{bmatrix}
\]
and assigning the input forces to \(\tau = V T \hat{\tau}\), we get a new input
matrix for \(\hat{\tau} \in \R^k\) given by \(\hat{B} = B V T = U \Sigma T\) 
which is still constant and full-rank. In particular, 
\(\hat{B}\tpose \hat{B} = T\tpose \Sigma\tpose \Sigma\tpose T = \Id{k}\).
\end{proof}

Let \(\mathbf{B} \in \R^{n\times n}\) be the following matrix:
\[
    \mathbf{B} = 
    \begin{bmatrix}
        B^\perp \\
        B\tpose \\
    \end{bmatrix}
\]
Since \(B^\perp\) is a left annihilator of \(B\) and both \(B^\perp\) and
\(B\tpose\) are right semi-orthogonal, it is easy to show that \(\mathbf{B}\) is
orthogonal:
\[
    \mathbf{B}\mathbf{B}\tpose = 
    \begin{bmatrix}
        B^\perp (B^\perp)\tpose & B^\perp B \\
        (B^\perp B)\tpose & B\tpose B
    \end{bmatrix} = \Id{n}
\]
Hence, \(\mathbf{B}\) is invertible with \(\mathbf{B}\inv = \mathbf{B}\tpose\).

The following theorem shows that \(\mathbf{B}\) provides a canonical
transformation into simply actuated coordinates, so that only the actuated momenta
are affected by the input forces.

\begin{thm}\label{thm:simply-actuated}
    Under Assumptions \ref{assm:B-const},\ref{assm:B-perp}, and
    \ref{assm:B-orthogonal}, the Hamiltonian system (\ref{eqn:hamiltonian}) is
    simply actuated with simply actuated coordinates 
    \(\{Q = \mathbf{B}q, P = \mathbf{B}p\}\). The resulting dynamics are 
    given by (\ref{eqn:simple-hamiltonian}),
    \begin{align}\label{eqn:simple-hamiltonian}
        \mathcal{H}(Q,P) &= 
        \frac{1}{2} P\tpose \Minv(Q) P + V(Q) \\
       &\begin{cases}
            \dot{Q} = \Minv(Q)P \\
            \dot{P} = -\frac{1}{2} (\Id{n} \otimes P\tpose) \nabla_Q \Minv(Q) P
                - \nabla_Q V(Q) + \simpleB \tau
        \end{cases} \nonumber
    \end{align}
    where
    \begin{align*}
        \Minv(Q) &:= \mathbf{B}D^{-1}(\mathbf{B}\tpose Q)\mathbf{B}\tpose \\
        V(Q) &:= P(\mathbf{B}\tpose Q) \\
    \end{align*}
\end{thm}
\vspace{-4.5em} % For some reason not having this puts a large space before proof
\begin{proof}
    The Poisson bracket between the functions \(f(q,p)\) and \(g(q,p)\) is defined by
    \cite{landau_mechanics} as follows:
    \[
        [f,g] := \sum \limits_{i=1}^n \pdiff{f}{p_i}\pdiff{g}{q_i} - 
                \pdiff{f}{q_i}\pdiff{g}{p_i}
    \]
    For any constant matrix \(A\), the transformation
    \(\{Q = Aq, P = Ap\}\) satisfies
    \(\pdiff{Q_i}{p_m} = \pdiff{P_i}{q_m} = 0\) for all 
    \(i,m \in \n\).
    Hence, for our new coordinates \(\{Q = \mathbf{B}q, P = \mathbf{B}p\}\),
    \begin{align*}
        [Q_i,Q_j] &:= \sum\limits_{m = 1}^n \pdiff{Q_i}{p_m}\pdiff{Q_j}{q_m} - 
        \pdiff{Q_i}{q_m}\pdiff{Q_j}{p_m} = 0 \\
        [P_i,P_j] &:= \sum\limits_{m=1}^n \pdiff{P_i}{p_m}\pdiff{P_j}{q_m} -
        \pdiff{P_i}{q_m}\pdiff{P_j}{p_m} = 0
    \end{align*}
    Since the matrix \(\mathbf{B}\) is orthogonal,
    \((\mathbf{B}_i)\tpose (\mathbf{B}\tpose)_j = (\mathbf{B}_i)\tpose (\mathbf{B}^{-1})_j = \delta_{i,j}\). Using this
    fact we see that the Poisson brackets between \(P_i\) and \(Q_j\) are given by:
    \begin{align*}
        [P_i,Q_j] &= \sum\limits_{m=1}^n\pdiff{P_i}{p_m}\pdiff{Q_j}{q_m}
        - \pdiff{P_i}{q_m}\pdiff{Q_j}{p_m} \\
                  &= \sum\limits_{m=1}^n \mathbf{B}_{i,m}\mathbf{B}_{j,m} - 0 \\
                  &= \sum\limits_{m=1}^n \mathbf{B}_{i,m}\mathbf{B}\tpose_{m,j} \\
                  &= (\mathbf{B}_i)\tpose \mathbf{B}\tpose_j \\
                  &= \delta_{i,j}
    \end{align*}

    Therefore, by (45.10) in \cite{landau_mechanics}, the coordinate change 
    \((Q = \mathbf{B}q, P = \mathbf{B}p)\) is a canonical transformation.
    Furthermore, since \(\dot{P} = \mathbf{B}\dot{p}\), the new input matrix is
    given by 
    \[
        \mathbf{B}B = \begin{bmatrix}
            B^\perp B \\
            B\tpose B \\
        \end{bmatrix} = 
        \begin{bmatrix}
            \Zmat{(n-k)\times k} \\
            \Id{k}
        \end{bmatrix}
    \]
    so the coordinates \(\{Q = (q_u,q_a), P = (p_u,p_a)\}\) are simply actuated coordinates for 
    \(\mathcal{H}\) as desired.
\end{proof}

@@

The downside to VHCs is that they do not use the additional information imparted
by the generalized velocities of a system.
For the child on a swing, whether they extend or retract their legs
depends on their direction of motion. This inherently requires knowledge of
their velocity, which precludes the usage of VHCs. 
Many authors have attempted to extend the theory of VHCs to enforce relations 
\(h(q,\dot{q}) = 0\) of the full state to account for this drawback. 
Since these relations use actuators to
restrict both the configuration and generalized velocity, they are called
virtual \textit{nonholonomic} constraints. This idea has been used for
human-robot interaction
\cite{vnhc_human_robot_cooperation,psd_based_vnhc_redundant_manipulator,haptic_vnhc},
error-reduction on time-delayed systems \cite{vnhc_time_delay_teleop},
and has shown marked improvements to the field of bipedal locomotion 
\cite{nhvc_dynamic_walking,
hybrid_zero_dynamics_bipedal_nhvcs,output_nhvc_bipedal_control}.
Most interestingly, the nonholonomic approach has been shown to be more robust
than standard VHCs when applied to bipedal robotics \cite{nhvc_incline_walking}.
This suggests that virtual nonholonomic constraints are more capable of
stabilizing specific energy levels while rejecting disturbances, all while
producing realistic biological motion. It is for this reason we choose to study
nonholonomic constraints in this thesis.

@@

    Moving on to \(\diff{}{t}\left\{dh_{p_u}\right\}\), a similar approach as

@@

% d/dt(1/2 * pudmat)
    For the next piece, observe that the \(i^\text{th}\) row of

@@

    In matrix form, this becomes:
